---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

# plotKML
plotKML: Visualization of Spatial and Spatio-Temporal Objects in Google Earth

<!-- badges: start -->

[![Build Status](https://travis-ci.org/Envirometrix/plotKML.svg?branch=master)](https://travis-ci.org/Envirometrix/plotKML)
[![R build status](https://github.com/Envirometrix/plotKML/workflows/R-CMD-check/badge.svg)](https://github.com/Envirometrix/plotKML/actions)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/plotKML)](https://cran.r-project.org/package=plotKML)
[![Github_Status_Badge](https://img.shields.io/badge/Github-0.7--1-blue.svg)](https://github.com/Envirometrix/plotKML)
<!-- badges: end -->

How to cite:

- Hengl, T., Roudier, P., Beaudette, D., & Pebesma, E. (2015). **plotKML: Scientific visualization of spatio-temporal data**. Journal of Statistical Software, 63(5), 1-25. https://www.jstatsoft.org/article/view/v063i05

## Installing

Install development versions from github:

```{r, eval = FALSE}
library(devtools)
install_github("envirometrix/plotKML")
```

## Integration with sf

I will now present new `sf` functions and compare with current `sp` functions. Start with POINTS

```{r}
suppressPackageStartupMessages({
  library(plotKML)
  library(sp)
  library(rgdal)
  library(sf)
})
# Load data
data(eberg)
coordinates(eberg) <- ~ X + Y
proj4string(eberg) <- CRS("+init=epsg:31467")
## subset to 20 percent:
eberg <- eberg[runif(nrow(eberg)) < .1, ]
# sp methods
plotKML(eberg["CLYMHT_A"], open.kml = FALSE)
plotKML(eberg["CLYMHT_A"], colour_scale = rep("#FFFF00", 2), points_names = "", open.kml = FALSE)

# convert eberg to sf format
eberg_sf <- st_as_sf(eberg)

# and apply sf methods
plotKML(eberg_sf["CLYMHT_A"], open.kml = FALSE)
plotKML(eberg_sf["CLYMHT_A"], colour_scale = rep("#FFFF00", 2), points_names = "", open.kml = FALSE)
```

Recent version of sf could show a warning message relative to the old PROJ4string specified by `get("ref_CRS", envir = plotKML.opts)`. 
Maybe `"ref_CRS"` should be changed or updated. 

We can now compare the two implementations: 

```{r}
all.equal(readLines("eberg__CLYMHT_A__.kml"), readLines("eberg_sf__CLYMHT_A__.kml"))
```

I think the warning messages are not important here. . 
The two string mismatches are simply caused by the different names and classes: 

```{r, warning = FALSE}
id_mismatches <- readLines("eberg__CLYMHT_A__.kml") != readLines("eberg_sf__CLYMHT_A__.kml")
readLines("eberg__CLYMHT_A__.kml")[id_mismatches]
readLines("eberg_sf__CLYMHT_A__.kml")[id_mismatches]
```

I don't know if that's a problem, but, here, both approaches return an error: 

```{r, error = TRUE}
plotKML(eberg, colour = CLYMHT_A, open.kml = FALSE)
plotKML(eberg_sf, colour = CLYMHT_A, open.kml = FALSE)
```

I can also use `kml_layer` functions: 

```{r}
data(eberg_grid)
gridded(eberg_grid) <- ~ x + y
proj4string(eberg_grid) <- CRS("+init=epsg:31467")
eberg_grid_sf <- st_as_sf(eberg_grid)

# sfc objects
kml_open("eberg_grids.kml")
kml_layer(st_geometry(eberg_grid_sf))
kml_close("eberg_grids.kml")

# sf objects
kml_open("eberg_grid_sf.kml")
kml_layer(eberg_grid_sf, colour = DEMSRT6, colour_scale = R_pal[["terrain_colors"]])
kml_layer(eberg_grid_sf, colour = TWISRT6, colour_scale = SAGA_pal[[1]])
kml_close("eberg_grid_sf.kml")
```

Then I can present `sf` functions for MULTIPOINT objects. 
The idea is exactly the same, but MULTIPOINT object are converted to POINT. 

```{r}
eberg_sf_MULTIPOINT <- eberg_sf %>% 
  dplyr::mutate(random_ID = sample(1:4, size = dplyr::n(), replace = TRUE)) %>% 
  dplyr::group_by(random_ID) %>% 
  dplyr::summarise()
plotKML(eberg_sf_MULTIPOINT["random_ID"], open.kml = FALSE)
```

Then I can work with LINESTRING, starting from `sp` example: 

```{r}
# sp
data(eberg_contours)
plotKML(eberg_contours, open.kml = FALSE)
plotKML(eberg_contours, colour = Z, altitude = Z, open.kml = FALSE)
# sf
eberg_contours_sf <- st_as_sf(eberg_contours)
plotKML(eberg_contours_sf, open.kml = FALSE)
plotKML(eberg_contours_sf, colour = Z, altitude = Z, open.kml = FALSE)
```

Again, the kml files are identical but for super small differences due to rounding errors: 

```{r, warning = FALSE}
all.equal(readLines("eberg_contours.kml"), readLines("eberg_contours_sf.kml"))

id_mismatches <- which(readLines("eberg_contours.kml") != readLines("eberg_contours_sf.kml"))
readLines("eberg_contours.kml")[id_mismatches[1:5]]
readLines("eberg_contours_sf.kml")[id_mismatches[1:5]]
```

The same ideas can be applied to MULTILINESTRING objects. 
The only problem is that  some of the features in `eberg_contous_sf` are not valid according to the simple feature definition of linestring. For example

```{r}
eberg_contours@lines[[4]]
```

since it would represent a LINESTRING with only 1 point: 

```{r, error = TRUE}
st_is_valid(st_geometry(eberg_contours_sf)[[4]], NA_on_exception = FALSE, reason = TRUE)
```

So I need to exclude these elements and create a MULTILINESTRING object, 

```{r}
ID_valid <- vapply(
  st_geometry(eberg_contours_sf), 
  function(x) isTRUE(st_is_valid(x)), 
  logical(1)
)
eberg_contous_sf_multi <- eberg_contours_sf %>% 
  dplyr::filter(ID_valid) %>% 
  dplyr::group_by(Z) %>% 
  dplyr::summarise()
plotKML(eberg_contous_sf_multi, colour = Z, altitude = Z, open.kml = FALSE)
```

I can also use `kml_layer` functions: 

```{r}
# sfc LINESTRING object
kml_open("eberg_contours_sfc.kml")
kml_layer(st_geometry(eberg_contours_sf))
kml_close("eberg_contours_sfc.kml")
```

Then I can work with POLYGON, starting from `sp` example: 

```{r}
# sp 
set.seed(1) # I set the seed to compare the kml files
data(eberg_zones)
plotKML(eberg_zones["ZONES"], open.kml = FALSE)
## add altitude:
zmin = 230
plotKML(eberg_zones["ZONES"], altitude = zmin + runif(length(eberg_zones)) * 500, open.kml = FALSE)

# sf objects with sfc_POLYGON geometry
eberg_zones_sf <- st_as_sf(eberg_zones)
set.seed(1)
plotKML(eberg_zones_sf["ZONES"], open.kml = FALSE)
plotKML(eberg_zones_sf["ZONES"], altitude = zmin + runif(length(eberg_zones)) * 500, open.kml = FALSE)
```

I can compare the results:  

```{r, warning = FALSE}
all.equal(readLines("eberg_zones__ZONES__.kml"), readLines("eberg_zones_sf__ZONES__.kml"))
```

The differences here are caused by extremely small rounding problems: 

```{r, warning = FALSE}
id_mismatches <- readLines("eberg_zones__ZONES__.kml") != readLines("eberg_zones_sf__ZONES__.kml")
readLines("eberg_zones__ZONES__.kml")[id_mismatches][1:5]
readLines("eberg_zones_sf__ZONES__.kml")[id_mismatches][42]
```

Again, the same ideas can be applied to MULTIPOLYGON objects: 

```{r}
eberg_zones_sf_MULTI <- eberg_zones_sf %>% 
  dplyr::group_by(ZONES) %>% 
  dplyr::summarise() %>% 
  st_cast("MULTIPOLYGON") 
plotKML(eberg_zones_sf_MULTI["ZONES"], open.kml = FALSE)
```

