---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

# plotKML
plotKML: Visualization of Spatial and Spatio-Temporal Objects in Google Earth

<!-- badges: start -->

[![Build Status](https://travis-ci.org/Envirometrix/plotKML.svg?branch=master)](https://travis-ci.org/Envirometrix/plotKML)
[![R build status](https://github.com/Envirometrix/plotKML/workflows/R-CMD-check/badge.svg)](https://github.com/Envirometrix/plotKML/actions)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/plotKML)](https://cran.r-project.org/package=plotKML)
[![Github_Status_Badge](https://img.shields.io/badge/Github-0.7--1-blue.svg)](https://github.com/Envirometrix/plotKML)
<!-- badges: end -->

How to cite:

- Hengl, T., Roudier, P., Beaudette, D., & Pebesma, E. (2015). **plotKML: Scientific visualization of spatio-temporal data**. Journal of Statistical Software, 63(5), 1-25. https://www.jstatsoft.org/article/view/v063i05

## Installing

Install development versions from github:

```{r, eval = FALSE}
library(devtools)
install_github("envirometrix/plotKML")
```

## Integration with sf

We will now present new `sf` functions and compare the with current `sp` approach. Start with POINTS

```{r}
suppressPackageStartupMessages({
  library(plotKML)
  library(sp)
  library(rgdal)
  library(sf)
})
# Load data
data(eberg)
coordinates(eberg) <- ~ X + Y
proj4string(eberg) <- CRS("+init=epsg:31467")
## subset to 20 percent:
eberg <- eberg[runif(nrow(eberg)) < .1, ]
# sp methods
plotKML(eberg["CLYMHT_A"], open.kml = FALSE)
plotKML(eberg["CLYMHT_A"], colour_scale = rep("#FFFF00", 2), points_names = "", open.kml = FALSE)

# convert eberg to sf format
eberg_sf <- st_as_sf(eberg)

# and apply sf methods
plotKML(eberg_sf["CLYMHT_A"], open.kml = FALSE)
plotKML(eberg_sf["CLYMHT_A"], colour_scale = rep("#FFFF00", 2), points_names = "", open.kml = FALSE)
```

Recent version of sf could show a warning message relative to the old PROJ4string specified by `get("ref_CRS", envir = plotKML.opts)`. 
Maybe `"ref_CRS"` should be changed or updated. 

We can now compare the two implementations: 

```{r}
all.equal(readLines("eberg__CLYMHT_A__.kml"), readLines("eberg_sf__CLYMHT_A__.kml"))
```

I think the warning messages are not important here. . 
The two string mismatches are simply caused by the different names and classes: 

```{r, warning = FALSE}
id_mismatches <- readLines("eberg__CLYMHT_A__.kml") != readLines("eberg_sf__CLYMHT_A__.kml")
readLines("eberg__CLYMHT_A__.kml")[id_mismatches]
readLines("eberg_sf__CLYMHT_A__.kml")[id_mismatches]
```

I don't know if that's a problem, but, here, both approaches return an error: 

```{r, error = TRUE}
plotKML(eberg, colour = CLYMHT_A, open.kml = FALSE)
plotKML(eberg_sf, colour = CLYMHT_A, open.kml = FALSE)
```

I can also use `kml_layer` functions: 

```{r}
data(eberg_grid)
gridded(eberg_grid) <- ~ x + y
proj4string(eberg_grid) <- CRS("+init=epsg:31467")
eberg_grid_sf <- st_as_sf(eberg_grid)

# sfc objects
kml_open("eberg_grids.kml")
kml_layer(st_geometry(eberg_grid_sf))
kml_close("eberg_grids.kml")

# sf objects
kml_open("eberg_grid_sf.kml")
kml_layer(eberg_grid_sf, colour = DEMSRT6, colour_scale = R_pal[["terrain_colors"]])
kml_layer(eberg_grid_sf, colour = TWISRT6, colour_scale = SAGA_pal[[1]])
kml_close("eberg_grid_sf.kml")
```

A more complex example: 

```{r, eval = FALSE}
shape = "http://maps.google.com/mapfiles/kml/pal2/icon18.png" 
kml_file = "kml_file_point.kml"
legend_file = "kml_legend.png"
kml_open(kml_file)
kml_layer(eberg_sf["CLYMHT_A"], colour = CLYMHT_A, size = 1, points_names = "", shape = shape, alpha = 0.75)
kml_legend.bar(eberg_sf$CLYMHT_A, legend.file = legend_file, legend.pal = SAGA_pal[[1]])
kml_screen(image.file = legend_file)
kml_close(kml_file)
kml_View(kml_file)
```


Then I can present `sf` functions for MULTIPOINT objects. 
The idea is exactly the same, but MULTIPOINT object are converted to POINT. 

```{r}
eberg_sf_MULTIPOINT <- eberg_sf %>% 
  dplyr::mutate(random_ID = sample(1:4, size = dplyr::n(), replace = TRUE)) %>% 
  dplyr::group_by(random_ID) %>% 
  dplyr::summarise()
plotKML(eberg_sf_MULTIPOINT["random_ID"], open.kml = FALSE)
```

Then I can work with LINESTRING, starting from `sp` example: 

```{r}
# sp
data(eberg_contours)
plotKML(eberg_contours, open.kml = FALSE)
plotKML(eberg_contours, colour = Z, altitude = Z, open.kml = FALSE)
# sf
eberg_contours_sf <- st_as_sf(eberg_contours)
plotKML(eberg_contours_sf, open.kml = FALSE)
plotKML(eberg_contours_sf, colour = Z, altitude = Z, open.kml = FALSE)
```

Again, the kml files are identical but for super small differences due to rounding errors: 

```{r, warning = FALSE}
all.equal(readLines("eberg_contours.kml"), readLines("eberg_contours_sf.kml"))

id_mismatches <- which(readLines("eberg_contours.kml") != readLines("eberg_contours_sf.kml"))
readLines("eberg_contours.kml")[id_mismatches[1:5]]
readLines("eberg_contours_sf.kml")[id_mismatches[1:5]]
```

The same ideas can be applied to MULTILINESTRING objects. 
The only problem is that  some of the features in `eberg_contous_sf` are not valid according to the simple feature definition of linestring. For example

```{r}
eberg_contours@lines[[4]]
```

since it would represent a LINESTRING with only 1 point: 

```{r, error = TRUE}
st_is_valid(st_geometry(eberg_contours_sf)[[4]], NA_on_exception = FALSE, reason = TRUE)
```

So I need to exclude these elements and create a MULTILINESTRING object, 

```{r}
ID_valid <- vapply(
  st_geometry(eberg_contours_sf), 
  function(x) isTRUE(st_is_valid(x)), 
  logical(1)
)
eberg_contous_sf_multi <- eberg_contours_sf %>% 
  dplyr::filter(ID_valid) %>% 
  dplyr::group_by(Z) %>% 
  dplyr::summarise()
plotKML(eberg_contous_sf_multi, colour = Z, altitude = Z, open.kml = FALSE)
```

I can also use `kml_layer` functions: 

```{r}
# sfc LINESTRING object
kml_open("eberg_contours_sfc.kml")
kml_layer(st_geometry(eberg_contours_sf))
kml_close("eberg_contours_sfc.kml")
```

Then I can work with POLYGON, starting from `sp` example: 

```{r}
# sp 
set.seed(1) # I set the seed to compare the kml files
data(eberg_zones)
plotKML(eberg_zones["ZONES"], open.kml = FALSE)
## add altitude:
zmin = 230
plotKML(eberg_zones["ZONES"], altitude = zmin + runif(length(eberg_zones)) * 500, open.kml = FALSE)

# sf objects with sfc_POLYGON geometry
eberg_zones_sf <- st_as_sf(eberg_zones)
set.seed(1)
plotKML(eberg_zones_sf["ZONES"], open.kml = FALSE)
plotKML(eberg_zones_sf["ZONES"], altitude = zmin + runif(length(eberg_zones)) * 500, open.kml = FALSE)
```

I can compare the results:  

```{r, warning = FALSE}
all.equal(readLines("eberg_zones__ZONES__.kml"), readLines("eberg_zones_sf__ZONES__.kml"))
```

The differences here are caused by extremely small rounding problems: 

```{r, warning = FALSE}
id_mismatches <- readLines("eberg_zones__ZONES__.kml") != readLines("eberg_zones_sf__ZONES__.kml")
readLines("eberg_zones__ZONES__.kml")[id_mismatches][1:5]
readLines("eberg_zones_sf__ZONES__.kml")[id_mismatches][42]
```

Again, the same ideas can be applied to MULTIPOLYGON objects: 

```{r}
eberg_zones_sf_MULTI <- eberg_zones_sf %>% 
  dplyr::group_by(ZONES) %>% 
  dplyr::summarise() %>% 
  st_cast("MULTIPOLYGON") 
plotKML(eberg_zones_sf_MULTI["ZONES"], open.kml = FALSE)
```

I will now present `plotKML` function applied to `stars` objects. 
We extended the `plotKML()` function to `stars` objects representing Raster Data Cubes creating a wrapper to RasterLayer method:  

```{r}
library(stars)
# Start with a raster data cube
(g <- read_stars(system.file("external/test.grd", package = "raster")))
plotKML(g, open.kml = FALSE)
```

Default methods do not work if the third dimension in a `stars` object do not represent a Date or POSIXct object: 

```{r, error = TRUE}
# Test with another example: 
(L7 <- read_stars(system.file("tif/L7_ETMs.tif", package = "stars")))
plotKML(L7, open.kml = FALSE)

# It fails. The problems is that there is no method definition for plotKML function for objects of class RasterBrick. 

# The code behind as(x, "Raster") is defined here: 
# https://github.com/r-spatial/stars/blob/e0c8506d8e00413721a1dfe83cc7f3d45b16c912/R/raster.R#L116-L124
# and it says that if length(dim(x)) > 2 (i.e. there is an extra dimension other than x and y), then it creates a RasterBrick object. 
```

but we can still plot the raster associated to one of the layers in the RasterBrick with a little bit of extra work: 

```{r}
plotKML(raster::raster(as(L7, "Raster"), layer = 1), open.kml = FALSE)
```

We are working on defining an extension to `stars` object with a temporal dimension as a wrapper around `RasterBrickTimeSeries` methods. 

Let's focus now on Vector data cubes: 

```{r}
# Vector data cube
data(air, package = "spacetime")
d = st_dimensions(station = st_as_sfc(stations), time = dates)
(aq = st_as_stars(list(PM10 = air), dimensions = d))
(agg = aggregate(aq, "months", mean, na.rm = TRUE))
plotKML(agg, open.kml = FALSE) 
```

Unfortunately there is a known bug in `plotKML`/`spacetime` and the following example fails:  

```{r, error = TRUE}
# Spatial aggregation: 
(a = aggregate(aq, st_as_sf(DE_NUTS1), mean, na.rm = TRUE))
plotKML(a) # error

# The error is in as(obj, "STIDF"), i.e. the conversion between STFDF and STIDF
```

Moreover, all the examples fail if the third dimension is not a temporal object (since the methods we defined as wrappers around STFDF class): 

```{r, error = TRUE}
nc = read_sf(system.file("gpkg/nc.gpkg", package="sf"))
nc.df = st_set_geometry(nc, NULL)
mat = as.matrix(nc.df[c("BIR74", "SID74", "NWBIR74", "BIR79", "SID79", "NWBIR79")])
dim(mat) = c(county = 100, var = 3, year = 2) # make it a 3-dimensional array
dimnames(mat) = list(county = nc$NAME, var = c("BIR", "SID", "NWBIR"), year = c(1974, 1979))
(nc.st = st_as_stars(pop = mat))
(nc.geom <- st_set_dimensions(nc.st, 1, st_geometry(nc)))
plotKML(nc.geom)
```

We can somehow fix this problem converting the `stars` object into `sf` format: 

```{r}
# We can use the following approach
plotKML(st_as_sf(nc.geom), open.kml = FALSE)
```

In some cases, it is still possible to redefine a temporal dimension (but unfortunately the example fails for the same bug as the previous case): 

```{r, error = TRUE}
(nc.geom <- st_set_dimensions(nc.geom, 3, as.Date(c("1974-01-01","1979-01-01"))))
(nc.geom <- split(aperm(nc.geom, c(1,3,2))))
plotKML(nc.geom)
```

