---
output: github_document
---
<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

<!-- README.md is generated from README.Rmd. Please edit that file -->

# plotKML
plotKML: Visualization of Spatial and Spatio-Temporal Objects in Google Earth

<!-- badges: start -->

[![Build Status](https://travis-ci.org/Envirometrix/plotKML.svg?branch=master)](https://travis-ci.org/Envirometrix/plotKML)
[![R build status](https://github.com/Envirometrix/plotKML/workflows/R-CMD-check/badge.svg)](https://github.com/Envirometrix/plotKML/actions)
[![CRAN_Status_Badge](http://www.r-pkg.org/badges/version/plotKML)](https://cran.r-project.org/package=plotKML)
[![Github_Status_Badge](https://img.shields.io/badge/Github-0.7--1-blue.svg)](https://github.com/Envirometrix/plotKML)
<!-- badges: end -->

How to cite:

- Hengl, T., Roudier, P., Beaudette, D., & Pebesma, E. (2015). **plotKML: Scientific visualization of spatio-temporal data**. Journal of Statistical Software, 63(5), 1-25. https://www.jstatsoft.org/article/view/v063i05

## Installing

Install development versions from github:

```{r, eval = FALSE}
library(devtools)
install_github("envirometrix/plotKML")
```

## Integration with sf

I will now present new `sf` functions and compare with current `sp` functions. Start with POINTS

```{r}
suppressPackageStartupMessages({
  library(plotKML)
  library(sp)
  library(rgdal)
  library(sf)
})
# Load data
data(eberg)
coordinates(eberg) <- ~ X + Y
proj4string(eberg) <- CRS("+init=epsg:31467")
## subset to 20 percent:
eberg <- eberg[runif(nrow(eberg)) < .1, ]
# sp methods
plotKML(eberg["CLYMHT_A"], open.kml = FALSE)
plotKML(eberg["CLYMHT_A"], colour_scale = rep("#FFFF00", 2), points_names = "", open.kml = FALSE)

# convert eberg to sf format
eberg_sf <- st_as_sf(eberg)

# and apply sf methods
plotKML(eberg_sf["CLYMHT_A"], open.kml = FALSE)
plotKML(eberg_sf["CLYMHT_A"], colour_scale = rep("#FFFF00", 2), points_names = "", open.kml = FALSE)
```

Recent version of sf could show a warning message relative to the old PROJ4string specified by `get("ref_CRS", envir = plotKML.opts)`. 
Maybe `"ref_CRS"` should be changed or updated. 

We can now compare the two implementations: 

```{r}
all.equal(readLines("eberg__CLYMHT_A__.kml"), readLines("eberg_sf__CLYMHT_A__.kml"))
```

I think the warning messages are not important here. . 
The two string mismatches are simply caused by the different names and classes: 

```{r, warning = FALSE}
id_mismatches <- readLines("eberg__CLYMHT_A__.kml") != readLines("eberg_sf__CLYMHT_A__.kml")
readLines("eberg__CLYMHT_A__.kml")[id_mismatches]
readLines("eberg_sf__CLYMHT_A__.kml")[id_mismatches]
```

I don't know if that's a problem, but, here, both approaches return an error: 

```{r, error = TRUE}
plotKML(eberg, colour = CLYMHT_A, open.kml = FALSE)
plotKML(eberg_sf, colour = CLYMHT_A, open.kml = FALSE)
```

Then I can present `sf` functions for MULTIPOINT objects: 

```{r}
eberg_sf_MULTIPOINT <- eberg_sf %>% 
  dplyr::mutate(random_ID = sample(1:4, size = dplyr::n(), replace = TRUE)) %>% 
  dplyr::group_by(random_ID) %>% 
  dplyr::summarise()
plotKML(eberg_sf_MULTIPOINT["random_ID"], open.kml = FALSE)
```

Then I can work with LINESTRINGS: 

```{r}
# sp
data(eberg_contours)
plotKML(eberg_contours, open.kml = FALSE)
plotKML(eberg_contours, colour = Z, altitude = Z, open.kml = FALSE)
# sf
eberg_contours_sf <- st_as_sf(eberg_contours)
plotKML(eberg_contours_sf, open.kml = FALSE)
plotKML(eberg_contours_sf, colour = Z, altitude = Z, open.kml = FALSE)
```

Again, the kml files are identical but for super small differences due to rounding errors: 

```{r, warning = FALSE}
all.equal(readLines("eberg_contours.kml"), readLines("eberg_contours_sf.kml"))

id_mismatches <- which(readLines("eberg_contours.kml") != readLines("eberg_contours_sf.kml"))
readLines("eberg_contours.kml")[id_mismatches[1:5]]
readLines("eberg_contours_sf.kml")[id_mismatches[1:5]]
````


